Sequential rank agreement methods for comparison of ranked lists

### 2016-08-04 latex/superranker-manuscript.tex
Small changes 2

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete or
  censored. The approach can identify change-points in the agreement
  of the lists, and the sequential changes of agreement as a function
  of the depth of the lists can be compared graphically to a
  permutation based reference set or evaluated using analytical asymptotic
  results.  The usefulness of these tools are illustrated using gene
  rankings, and using data from two Danish ovarian cancer studies
  where we assess the within and between agreement of different
  statistical classification methods.
\end{abstract}

@@

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to
compare multiple lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, it places
higher weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
change-point, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide asymptotical and randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

@@

The simplest case occurs when all $L$ lists are fully observed, \ie,
we have observed the rank of all $P$ items for all $L$ lists. Fully
observed ranked lists are common and arise, for example, when
different statistical analysis methods are applied to a single dataset
to produce lists of predictors ranked according to their importance,
or if the same analysis method is applied to data from different
populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal unequal variances two-sample $t$ tests,
marginal logistic regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to minimum $p$ value, for
logistic regression the genes were ordered by size of the
corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}.

The sequential rank agreement curve plotted in Figure~\ref{fig:example1}
roughly shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that the sequential rank agreement limits are better towards the top
of the lists (smaller values on the $y$ axis) corresponding to
\emph{better} agreement than towards the tails of the lists. It is clear
from the curve in Figure~\ref{fig:example1} that there is a
substantial drop in agreement even after the first depth. Thus, if we
were to restrict attention to a small set of predictors then we would
focus on predictors 2124, 829, and 378 as seen in Table~\ref{tab1}.

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & T & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data /black
   line). Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available and the rest
   of the items have been censored. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

 If the lists agree on all items then the sequential rank agreement is
 zero for all depths. Generally, if there are changes in the level of
 rank agreement then this suggests that there are sets of items that
 are ranked similarly in all lists and other sets of items that have
 been assigned vastly different ranks in the lists. In gene
 association studies we generally expect the agreement among the lists
 to be better towards the top of the lists and worse towards the end
 of the lists. In this case the sequential rank agreement curve will
 start at a low level and then increase until it levels off as seen in
 Figure~\ref{fig:example1}.

\subsection{Analysis of partial/censored lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression based on the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ (\ie, top $d$ lists) such that the
measurement of agreement is restricted to the first $d$ elements of
each list, or when the analyst only wants to consider the ranking of
items that have been found to be statistically significant.

@@

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(e.g., the mean of the unassigned ranks) to the censored items since
that would result in \emph{less} variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are censored in multiple
lists.

@@

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For partially ranked lists we also
expect a change-point around the depth where the lists are
censored. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

@@

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

To obtain the distribution under $\widetilde H_0$ the idea is to
repeat the ranking procedures for unassociated data many times.  Thus,
we first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

@@

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. However, the
censored data also suggests that there may be at most 1 or 2 ranked
items towards the top of the lists that yield a result better than
what would be expected (the bottom-right plot).

It is important to stress that neither $H_0$ nor $\widetilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the data set. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

\subsection{Asymptotic inference of change in agreement}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depth which
satisfies a change point criterion since that corresponds to a change
in agreement among the list ranks. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As in \cite{hall:schi:2012} we consider an infinite set
of lists and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to change with $L$
because in most applications the lengths of the lists are fixed.

@@

We now define the change point as the first crossing point of the
sequential rank agreement and a threshold function
$q\colon\,\{1,\ldots,P\} \mapsto \mathbb{R}_{\geq 0}$. The values of
$q$ could be a deterministic constant, or in the experimental setting,
for example, the limits-of-agreement obtained in randomly permuted
lists corresponding to the null-hypothesis in equation
(\ref{eq:permutationHypothesis}). We define the superlevel set of the
sequential rank agreement with respect to $q$ as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}.
\end{align}
A change-point $d^\ast(q)$ in the list agreement is then defined by the position
\begin{equation}
  d^\ast(q) =
  \begin{cases}
    \inf(\mathcal{L}(q)) & |\mathcal{L}(q)| > 0\\
  P & |\mathcal{L}(q)| = 0
  \end{cases}
\end{equation}
corresponding to the first list depth where the sequential rank
agreement exceeds the threshold if such a position exists. Otherwise,
the full list is in agreement according to $q$ and the change-point is
set to the length of the lists. The empirical superlevel set is
similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where we allow the threshold function to depend on the sample size as
well. The estimated change-point is therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P.
\end{align}

The consistency of the estimated change-point, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}\label{thecor}
Let $\widehat{q}_L$ be a positive threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then, $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\begin{proof}
See Appendix \ref{sec:appB}.
\end{proof}
\end{corollary}

@@

\section{Application to ovarian cancer data}
We now consider an application of the sequential rank agreement to two
data sets consisting of MALDI-TOF (Matrix-Assisted Laser
Desorption/Ionization Time Of Flight) mass spectra obtained from blood
samples from patients with either benign or malignant ovarian
tumors. The data sets are sub-samples of the Danish MALOVA and DACOVA
study populations.

@@

Each spectrum consists of $49642$ samples over a range of mass-to-charge ratios
between $800$ to $20000$ Dalton which we downsample on an equidistant grid of
5000 points by linear interpolation. We then preprocess the downsampled
spectra individually by first removing the slow-varying baseline intensity
with the SNIP algorithm \citep{ryan1988snip} followed by a
normalization with respect to the total ion count. Finally, we standardize
the $5000$ predictors to have column-wise zero mean and unit variance in
each data set.

We use the two data sets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of four different statistical
classification methods in how they rank the predictors according to
their importance for distinguishing benign and malignant tumors. In
the second scenario we assess the agreement among rankings of
individual predicted risks of having a malignant tumor. The first
scenario is relevant in the context of biomarker discovery and the
latter is important e.g., when ranking patients according to immediacy of
treatment.

@@

For most of the sequential rank agreement curves we see, as expected,
that they start low, indicating good agreement, followed by an
increase until they approximately become constant. This has the
interpretation that the agreement across the different sub-samples is
higher in the top as compared to the tail of the lists for all these
classification methods. The change-points where the curves become
approximately constant are the list depths where the ranks of the
remaining items become close to uniformly random.

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient values
below a given threshold thereby introducing an artificial censoring of the lists.
For the Ridge models tuned with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated
using Algorithm \ref{sra-algorithm} with $B=1000$ and $P=5000$. The corresponding
curve from Figure \ref{fig:app1} (right panel) is shown for comparison.
Even though the number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing.

@@

\subsection{Agreement of individual risk predictions}
\label{sec:airp}
To assess the stability of the individual risk predictions we apply
the predictors from the DACOVA data set to each of the models. The
predicted probabilities are then ranked in decreasing order such that the
patients with the highest risk of a malignant tumor appears in the top
of the list. Figure \ref{fig:app1} (left) shows the sequential rank
agreement separately for each method, based on the 1000 risk predictions
obtained from the models trained in the same $1000$ random sub-samples
of the MALOVA data.

@@

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreement curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wise $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a change-point analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete or
  censored. The approach can identify changepoints in the agreement
  of the lists, and the sequential changes of agreement as a function
  of the depth of the lists can be compared graphically to a
  permutation based reference set or evaluated using analytical asymptotic
  results.  The usefulness of these tools are illustrated using gene
  rankings, and using data from two Danish ovarian cancer studies
  where we assess the within and between agreement of different
  statistical classification methods.
\end{abstract}

@@

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to
compare multiple lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, it places
higher weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
changepoint, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide asymptotical and randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

@@

The simplest case occurs when all $L$ lists are fully observed, \ie, when
we observe the rank of all $P$ items for all $L$ lists. Fully
observed ranked lists are common and arise, for example, when
different statistical analysis methods are applied to a single dataset
to produce lists of predictors ranked according to their importance,
or if the same analysis method is applied to data from different
populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in the top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal unequal variances two-sample $t$ tests,
marginal logistic regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to minimum $p$ value, for
logistic regression the genes were ordered by size of the
corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}.

The sequential rank agreement curve plotted in Figure~\ref{fig:example1}
roughly shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that the sequential rank agreement limits are better towards the top
of the lists (smaller values on the $y$ axis corresponds to
\emph{better} agreement) than towards the tails of the lists. It is clear
from the curve in Figure~\ref{fig:example1} that there is a
substantial drop in agreement even after the first depth. Thus, if we
were to restrict attention to a small set of predictors then we would
focus on predictors 2124, 829, and 378 as seen in Table~\ref{tab1}.

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & Welsh's $t$ & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data (black
   line). Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available and the rest
   of the items have been censored in each list. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

% If the lists agree on all items then the sequential rank agreement is
% zero for all depths. 
 Generally, if there are changes in the level of
 rank agreement then this suggests that there are sets of items that
 are ranked similarly in all lists and other sets of items that have
 been assigned vastly different ranks in the lists. For example, in gene
 association studies we generally expect the agreement among the lists
 to be better towards the top of the lists and worse towards the end
 of the lists. In this case the sequential rank agreement curve will
 start at a low level and then increase until it levels off as seen in
 Figure~\ref{fig:example1}.

\subsection{Analysis of partial/censored lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression based on the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur if the analyst decides
to censor all lists at depth $d$ (\ie, top $d$ lists) such that the
measurement of agreement is restricted to the first $d$ elements of
each list, or when the analyst only wants to consider the ranking of
items that have been found to be statistically significant.

@@

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(\eg, the mean of the unassigned ranks) to the censored items since
that would result in \emph{less} variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are censored in multiple
lists.

@@

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For partially ranked lists we also
expect a changepoint around the depth where the lists are
censored. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

@@

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

To obtain the distribution under $\widetilde H_0$ the idea is to
repeat the ranking procedures for unassociated data many times.  Thus,
we first permute the outcome variable of the dataset on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
dataset to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. This randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

@@

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. The censored
data also suggests that there may be at most 1 or 2 ranked items
towards the top of the lists that yield a result better than what
would be expected (the bottom-right plot).

It is important to stress that neither $H_0$ nor $\widetilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the dataset. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

\subsection{Asymptotic inference of change in agreement}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depth which
satisfies a changepoint criterion since that corresponds to a change
in agreement among the list ranks. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As in \cite{hall:schi:2012} we consider an infinite set
of lists and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to change with $L$
because in most applications the lengths of the lists are fixed.

@@

We now define the changepoint as the first crossing point of the
sequential rank agreement and a threshold function
$q\colon\,\{1,\ldots,P\} \mapsto \mathbb{R}_{\geq 0}$. The values of
$q$ could be a deterministic constant or, % in the experimental setting,
for example, the limits-of-agreement obtained in randomly permuted
lists corresponding to the null-hypothesis in equation
(\ref{eq:permutationHypothesis}). We define the superlevel set of the
sequential rank agreement with respect to $q$ as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}.
\end{align}
A changepoint $d^\ast(q)$ in the list agreement is then defined by the position
\begin{equation}
  d^\ast(q) =
  \begin{cases}
    \inf(\mathcal{L}(q)) & |\mathcal{L}(q)| > 0\\
  P & |\mathcal{L}(q)| = 0
  \end{cases}
\end{equation}
corresponding to the first list depth where the sequential rank
agreement exceeds the threshold if such a position exists. Otherwise,
the full list is in agreement according to $q$ and the changepoint is
set to the full length of the lists. The empirical superlevel set is
similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where we allow the threshold function to depend on the sample size as
well. The estimated changepoint is therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P.
\end{align}

The consistency of the estimated changepoint, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}\label{thecor}
Let $\widehat{q}_L$ be a positive threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then, $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\begin{proof}
See Appendix \ref{sec:appB}.
\end{proof}
\end{corollary}

@@

\section{Application to ovarian cancer data}
We now consider an application of the sequential rank agreement to two
datasets consisting of MALDI-TOF (Matrix-Assisted Laser
Desorption/Ionization Time Of Flight) mass spectra obtained from blood
samples from patients with either benign or malignant ovarian
tumors. The datasets are sub-samples of the Danish MALOVA and DACOVA
study populations.

@@

Each spectrum consists of $49642$ samples over a range of mass-to-charge ratios
between $800$ to $20000$ Dalton which we downsample on an equidistant grid of
5000 points by linear interpolation. We then preprocess the downsampled
spectra individually by first removing the slow-varying baseline intensity
with the SNIP algorithm \citep{ryan1988snip} followed by a
normalization with respect to the total ion count. Finally, we standardize
the $5000$ predictors to have column-wise zero mean and unit variance in
each dataset.

We use the two datasets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of four different statistical
classification methods in how they rank the predictors according to
their importance for distinguishing benign and malignant tumors. In
the second scenario we assess the agreement among rankings of
individual predicted risks of having a malignant tumor. The first
scenario is relevant in the context of biomarker discovery and the
latter is important e.g., when ranking patients according to immediacy of
treatment.

@@

For most of the sequential rank agreement curves we see, as expected,
that they start low, indicating good agreement, followed by an
increase until they approximately become constant. This has the
interpretation that the agreement across the different sub-samples is
higher in the top as compared to the tail of the lists for all these
classification methods. The changepoints where the curves become
approximately constant are the list depths where the ranks of the
remaining items become close to uniformly random.

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this dataset and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient values
below a given threshold thereby introducing an artificial censoring of the lists.
For the Ridge models tuned with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated
using Algorithm \ref{sra-algorithm} with $B=1000$ and $P=5000$. The corresponding
curve from Figure \ref{fig:app1} (right panel) is shown for comparison.
Even though the number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing.

@@

\subsection{Agreement of individual risk predictions}
\label{sec:airp}
To assess the stability of the individual risk predictions we apply
the predictors from the DACOVA dataset to each of the models. The
predicted probabilities are then ranked in decreasing order such that the
patients with the highest risk of a malignant tumor appears in the top
of the list. Figure \ref{fig:app1} (left) shows the sequential rank
agreement separately for each method, based on the 1000 risk predictions
obtained from the models trained in the same $1000$ random sub-samples
of the MALOVA data.

@@

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreement curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wise $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a changepoint analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.
